Values disclosures and trust in science: A replication study

### 2022-08-11 paper/paper.Qmd
ready to submit to Frontiers

---
title: "Values disclosures and trust in science: A replication study"
author: "Daniel J. Hicks and Emilio Lobato"
abstract: |
  While philosophers of science generally agree that social, political, and ethical values can play legitimate roles in science, there is active debate over whether scientists should disclosure such values in their public communications.  This debate depends, in part, on empirical claims about whether values disclosures might undermine public trust in science.  In a previous study, Elliott et al. used an online experiment to test this empirical claim.  The current paper reports a replication attempt of their experiment. Comparing results of the original study and our replication, we do not find evidence for a transparency penalty or "shared values" effect, but do find evidence that the content of scientific conclusions (whether or not a chemical is found to cause harm) might effect perceived trustworthiness and that scientists who value public health and disclose this value might be perceived as more trustworthy.  
toc: true
execute:
    echo: false
bibliography: |
  transparency-Hicks.yaml
format: 
    html: default
    pdf: default
always_allow_html: true
---

@@

![Participant values by political ideology. (A) Absolute counts, (B) shares within political ideology categories.  Panel (A) shows that our sample substantially over-represents political liberals.](`r here(out_dir, '03_part_values.png')`){#fig-part-values}

@@

![Directed acyclic graph (DAG) for analysis of the shared values effect.  If the DAG is faithful, adjusting for scientist values (`sci_values`) and participant values (`part_values`) is sufficient for estimating the effect of shared values on perceived trustworthiness (`METI`).  In particular, adjusting for demographics should not change the regression coefficient on shared values.](`r here(out_dir, '03_shared_values_dag.png')`){#fig-shared-dag}

@@

![Data, group means, and 95% confidence intervals for H5-consumer, interaction of consumer risk sensitivity and participant values. Panels correspond to participant values.](`r here(out_dir, '03_conclusion_part.png')`){#fig-conclusion-part}

@@

![Data, group means, and 95% confidence intervals for H5-shared, interaction of shared values and participant values.  Panels correspond to participant values.](`r here(out_dir, '03_shared_part.png')`){#fig-shared-part}

@@

![Data, group means, and 95% confidence intervals for a potential interaction of scientist values and participant values. Panels correspond to participant values. This figure is identical to @fig-shared-part, except the columns in the economic growth panel have been switched.](`r here(out_dir, '03_sci_part.png')`){#fig-sci-part}

@@

>>>>

---
title: "Values disclosures and trust in science: A replication study"
# author: 
#     - name: Daniel J. Hicks
#       orcid: 0000-0001-7945-4416
#       email: dhicks4@ucmerced.edu
#       affiliations:
#           - ref: ucm
#       attributes:
#           corresponding: true
#     - name: Emilio Lobato
#       orcid: 0000-0002-3066-2932
#       email: elobato@ucmerced.edu
#       affiliations: 
#         - ref: ucm

affiliations: 
    - id: ucm
      name: University of California, Merced
      address: 5200 N Lake Road
      city: Merced
      state: CA
      postal-code: 95343

abstract: |
  While philosophers of science generally agree that social, political, and ethical values can play legitimate roles in science, there is active debate over whether scientists should disclosure such values in their public communications.  This debate depends, in part, on empirical claims about whether values disclosures might undermine public trust in science.  In a previous study, Elliott et al. used an online experiment to test this empirical claim.  The current paper reports a replication attempt of their experiment. Comparing results of the original study and our replication, we do not find evidence for a transparency penalty or "shared values" effect, but do find evidence that the content of scientific conclusions (whether or not a chemical is found to cause harm) might effect perceived trustworthiness and that scientists who value public health and disclose this value might be perceived as more trustworthy.

toc: false
execute:
    echo: false
bibliography: |
  transparency-Hicks.yaml
format: 
    html: default
    pdf: default
always_allow_html: true
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
---

| Daniel J. Hicks^1\*^, Emilio Lobato^1^
| ^1^: University of California, Merced, 5200 N Lake Road, Merced, CA, 95343
| ^\*^: Address correspondence to <dhicks4@ucmerced.edu>
| 
| Keywords: trust, trust in science, values in science, replication, philosophy of science

@@

![Participant values by political ideology. (A) Absolute counts, (B) shares within political ideology categories.  Panel (A) shows that our sample substantially over-represents political liberals.](fig2_part_values.png){#fig-part-values}

@@

![Directed acyclic graph (DAG) for analysis of the shared values effect.  If the DAG is faithful, adjusting for scientist values (`sci_values`) and participant values (`part_values`) is sufficient for estimating the effect of shared values on perceived trustworthiness (`METI`).  In particular, adjusting for demographics should not change the regression coefficient on shared values.](fig3_shared_values_dag.png){#fig-shared-dag}

@@

![Data, group means, and 95% confidence intervals for H5-consumer, interaction of consumer risk sensitivity and participant values. Panels correspond to participant values.](fig4_conclusion_part.png){#fig-conclusion-part}

@@

![Data, group means, and 95% confidence intervals for H5-shared, interaction of shared values and participant values.  Panels correspond to participant values.](fig5_shared_part.png){#fig-shared-part}

@@

![Data, group means, and 95% confidence intervals for a potential interaction of scientist values and participant values. Panels correspond to participant values. This figure is identical to @fig-shared-part, except the columns in the economic growth panel have been switched.](fig6_sci_part.png){#fig-sci-part}

@@

# Acknowledgments #

For feedback on this project, thanks to Kevin Elliott and participants at the Values in Science conference and Political Philosophy and Values in Medicine, Science, and Technology 2022 conference.

# Funding #

Funding for this project was provided by the University of California, Merced.
